---
title: "Resting Heartrates (flipped assignment 2)"
author: "Stephen Ratcliff, Kameron Garza, Forest Kingfisher"
date: "`r Sys.Date()`"
output: html_document
---

In this analysis, we analyze the resting heart rate data collected on 65 random males and 65 random females. We read the data from a Comma Separated Value (.csv) file posted to the web consisting of three data values per participant:

* `Temp`: body temperature in Fahrenheit degrees), 
* `Sex`: biological gender, either male = 1 or female = 2 
* `Beats`: heart rate, in beats per minute


```{r}
#Import normtemp.csv from url
dat <- read.csv("https://raw.githubusercontent.com/tmatis12/datafiles/main/normtemp.csv")
```

### Descriptive Statistics

Summary statistics show that the mean and median averages are nearly equal,
indicating that our distrubution is symetric and there is a good chance we can 
use a normal distribution to approximate it. However, the first quartile is 
closer to the average than the third, indicating some left-skew in the 
distribution which may or may not be relevant. The minimum and maximum are 
within three standard deviations of the average (between 56 and 90), so we have 
no outliers to investigate. 


```{r}
malebeats = subset(dat,Sex==1)

# Descriptive Statistics
summary(malebeats$Beats)

# Male Heart Rate Quartiles
quantile(malebeats$Beats)

# Sample Standard Deviation
sd(malebeats$Beats)
```

### Histograms

Our histogram shows some non-normality in a mostly symmetric distribution, which 
we expect with a sample size of 65. Our density histogram confirms that the 
sample distribution is roughly symmetrical and has a slight left-skew.


```{r}
# Male Heart Rate Histogram
barplot(table(malebeats$Beats),xlab = "Heart Rate",ylab = "Frequency",col = "blue",main="Male Heart Rate")
plot(density(malebeats$Beats),col="blue",main="smoothed histogram of heartrates - males")
```

### Normal Probability Plot

Comparing the individual data values for heartrate to a theoretical average line 
(qqnorm connects central quartiles), the data trends similar to this theoretical 
linear approximation, apparent even without confidence intervals.

```{r}
# Male Heart Rate Normal Probability Plot
qqnorm(malebeats$Beats,col = 'blue', main="Male Heart Rate Normal Probability Plot", ylab = "Heartrates")
qqline(malebeats$Beats)
```

# Female Descriptive Statistics

Summary statistics show that the mean and median averages are slightly 
different, indicating that our distribution is not symetric. However, the 
first quartile is almost exactly equidistant from the mean as the third, 
indicating less skew in the overall distribution compared to the asymetry near 
the mean. This bears more examination in the histograms. The minimum and maximum 
are within three standard deviations of the mean (between 50 and 98) and 
median (between 52 and 100), so we have no outliers to investigate. 

```{r}
femalebeats = subset(dat,Sex==2)

# Descriptive Statistics
summary(femalebeats$Beats)

# Female Heart Rate Quantiles
quantile(femalebeats$Beats)

# Sample Standard Deviation
sd(femalebeats$Beats)
```

### Histograms

Our histogram shows a flat and wide normal-like distribution with a much more 
distinct left skew than the male heartrate histogram. Our density histogram 
confirms that the sample distribution is left-skewed but resembles the 
distinctive bell-curve shape of a normal distribution.


```{r}
# Female Heart Rate Histogram
barplot(table(femalebeats$Beats),xlab = "Heart Rate",ylab = "Frequency",col = "pink",main="Female Heart Rate")
plot(density(femalebeats$Beats),col="pink",main="smoothed histogram of heartrates - females")

```

### Normal Probability Plot

Comparing the individual data values for heartrate to a theoretical average line 
(qqnorm connects central quartiles), the data trends similar to this theoretical 
linear approximation, apparent even without confidence intervals.

```{r}
# Female Heart Rate Normal Probability Plot
qqnorm(femalebeats$Beats,col = 'pink', main="Female Heart Rate Normal Probability Plot", ylab = "Heartrates")
qqline(femalebeats$Beats)
```

# Male vs Female Box Plot Comparison

These side by side box plots that compare the resting heart rate of males and 
females illustrates the following comparisons in heartrate distributions:
 * both plots are skewed left, so above average heartrates for both genders show less variance than lower heartrates
 * The central quartiles for both genders clusters overlap, so average heartrates should be in similar ranges
 * The heartrate range, and consequently variance, was larger for females than males
 * The average heartrate was higher for females than males

```{r}
# boxplots comparing males' and females' heartrates
boxplot(Beats~Sex, data = dat, col=c("blue","pink"), names = c("Male", "Female"),xlab = "Gender",ylab = "Heart Rates",main="Heartrate - Male Vs Female")
```

# Complete Code

Here we display the complete R code used in this analysis.

```{r complete R code, eval = FALSE}
#Import normtemp.csv from url
dat <- read.csv("https://raw.githubusercontent.com/tmatis12/datafiles/main/normtemp.csv")

malebeats = subset(dat,Sex==1)

# Descriptive Statistics
summary(malebeats$Beats)

# Male Heart Rate Quartiles
quantile(malebeats$Beats)

# Sample Standard Deviation
sd(malebeats$Beats)

# Male Heart Rate Histogram
barplot(table(malebeats$Beats),xlab = "Heart Rate",ylab = "Frequency",col = "blue",main="Male Heart Rate")
plot(density(malebeats$Beats),col="blue",main="smoothed histogram of heartrates - males")

# Male Heart Rate Normal Probability Plot
qqnorm(malebeats$Beats,col = 'blue', main="Male Heart Rate Normal Probability Plot", ylab = "Heartrates")
qqline(malebeats$Beats)

femalebeats = subset(dat,Sex==2)

# Descriptive Statistics
summary(femalebeats$Beats)

# Female Heart Rate Quantiles
quantile(femalebeats$Beats)

# Sample Standard Deviation
sd(femalebeats$Beats)

# Female Heart Rate Histogram
barplot(table(femalebeats$Beats),xlab = "Heart Rate",ylab = "Frequency",col = "pink",main="Female Heart Rate")
plot(density(femalebeats$Beats),col="pink",main="smoothed histogram of heartrates - females")

# Female Heart Rate Normal Probability Plot
qqnorm(femalebeats$Beats,col = 'pink', main="Female Heart Rate Normal Probability Plot", ylab = "Heartrates")
qqline(femalebeats$Beats)

# boxplots comparing males' and females' heartrates
boxplot(Beats~Sex, data = dat, col=c("blue","pink"), names = c("Male", "Female"),xlab = "Gender",ylab = "Heart Rates",main="Heartrate - Male Vs Female")

```
